---
title: "Publication-Ready Tables of Particular Statistical Tests and Models with {gtsummary}"
description: |
  Find a review of incredibly useful {gtsummary} package in a separate blog-post. Here I'll just collect all the possible Statsitcal Tests and Models, {gtsummary} can help with.
author:
  - name: Yury Zablotski
    url: https://yuzar-blog.netlify.app/
date: "`r format(Sys.time(), '%B %d, %Y')`"
categories:
  - videos
  - statistics
  - models
  - machine learning
preview: thumbnail_gtsummary2.png
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 6
    code_download: true
bibliography: /Users/zablotski/Documents/library.bib
#csl: american-political-science-association.csl
biblio-style: apalike
link-citations: yes
linkcolor: blue
#draft: true
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
library(ggstatsplot)
library(tidyverse)
```

# Quick review of the package as a video

```{r, eval=T, echo=F}
vembedr::embed_youtube("hyP3Hx_1kTM") 
```




# Individual Statistical tests with p-values


## Wilcoxon rank sum test for independent samples



```{r}
library(gtsummary)
trial %>% 
  select(trt, age) %>% 
  tbl_summary(by = trt) %>% 
  add_p()
```

## Student's Two Sample t-test for independent samples with equal variances

```{r}
trial %>% 
  select(trt, age) %>% 
  tbl_summary(
    by = trt, 
    statistic = age ~ "{mean} ({sd})",
    missing = "no") %>% 
  add_n() %>% 
  add_p(age ~ "t.test", test.args = age ~ list(var.equal = TRUE))
```





## Welch Two Sample t-test for independent samples with unequal variances

```{r}
trial %>% 
  select(trt, age) %>% 
    tbl_summary(
    by = trt, 
    statistic = age ~ "{mean} ({sd})") %>%
  add_p(age ~ "t.test")
```

## Paired t-test

```{r}
# install.packages("BSDA")       # for Fitness data
library(BSDA)

Fitness %>%
  tbl_summary(
    by = test, 
    include = -subject,
    statistic = number ~ "{mean} ({sd})") %>%
  add_p(test = number ~ "paired.t.test", group = subject)

# prove
# make wide format
d <- Fitness %>% 
  pivot_wider(
    id_cols     = subject, 
    names_from  = test, 
    values_from = number) %>% 
  mutate(difference =  After - Before)

t.test(d$Before, d$After, paired = T)

ggwithinstats(
  data = Fitness,
  x    = test, 
  y    = number, 
  type = "parametric"
)
```


## Paired Wilcoxon test

```{r}
# install.packages("BSDA")       # for Fitness data
library(BSDA)

Fitness %>%
  tbl_summary(
    by = test, 
    include = -subject) %>%
    add_p(test = number ~ "paired.wilcox.test", group = subject, test.args = number ~ list(exact = F))

# prove
wilcox.test(d$Before, d$After, paired = T, exact = F)

ggwithinstats(
  data = Fitness,
  x    = test, 
  y    = number, 
  type = "np"
)
```



## Kruskal-wallis test

```{r}
trial %>% 
  select(age, stage) %>% 
  tbl_summary(by = stage) %>% 
  add_p()

ggbetweenstats(trial, stage, age, type = "np")
```

## Fisher's ANOVA

```{r}
trial %>% 
  select(age, stage) %>% 
  tbl_summary(
    by = stage, 
    statistic = age ~ "mean ({sd})") %>% 
  add_p(
    age ~ "aov", 
    pvalue_fun = ~ style_pvalue(.x, digits = 2),
    test.args = age ~ list(var.equal = T))

ggbetweenstats(trial, stage, age, var.equal = T)
```


## Welsh's ANOVA (not implemented yet: Nov. 2022)

```{r}
trial %>% 
  select(age, stage) %>% 
  tbl_summary(
    by = stage, 
    statistic = age ~ "mean ({sd})") %>% 
  add_p(
    age ~ "aov", 
    pvalue_fun = ~ style_pvalue(.x, digits = 2),
    test.args = age ~ list(var.equal = F))

ggbetweenstats(trial, stage, age, var.equal = F)
```

## Repeated Measures ANOVA (not implemented yet: Nov. 2022)


## Friedman test (not implemented yet: Nov. 2022)


## Pearson's Chi-squared test

```{r}
trial %>% 
  select(trt, stage) %>% 
  tbl_summary(by = stage) %>% 
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 3))

ggbarstats(trial, trt, stage, label = "both")
```


## Fisher's test

If number of observations in the categorical variable are <5, it takes Fisher's test automatically, which is kind of genius. 

```{r}
mtcars %>% 
  select(am, cyl) %>% 
  tbl_summary(by = am) %>% 
  add_p()
```


But we can force the function calculate Fisher's test with >5 numbers too.


```{r}
trial %>% 
  select(trt, stage) %>%
  tbl_summary(by = trt) %>% 
  add_p(
    stage ~ "fisher.test", 
    pvalue_fun = ~ style_pvalue(.x, digits = 2))
```




## McNemar's Chi-squared test with continuity correction

### short table format

```{r}
set.seed(9) # for reproducibility 

data2 <- data.frame(
  id     = c(1:20),
  before = sample(c("Yes", "No", "Yes", "Yes"), 20, replace = TRUE),
  after  = sample(c("No", "Yes", "No", "No"), 20, replace = TRUE))


data2 %>% 
  select(-id) %>% 
  tbl_cross(row = before, col = after) %>% 
  add_p("mcnemar.test.wide")

mcnemar.test(x = data2$before, y = data2$after, correct = T)
```

### long table format

```{r}
data <- data2 %>% 
  pivot_longer(cols = 2:3, names_to = "time", values_to = "result")

data %>%
  tbl_summary(
    by = time, 
    include = result,
    type = list(result ~ "categorical")) %>%
  add_p(
    everything() ~ "mcnemar.test", 
    group = id,
    pvalue_fun = ~style_pvalue(., digits = 3)
  ) 
```


## McNemar's Chi-squared test without continuity correction

```{r}
data2 %>% 
  select(-id) %>% 
  tbl_cross(row = before, col = after) %>% 
  add_p("mcnemar.test.wide", test.args=list(correct = F))

mcnemar.test(x = data2$before, y = data2$after, correct = F)

ggbarstats(
  data = data2, 
  x = before,
  y = after, 
  paired = T
)
```




## Proportion test: 2-sample test for equality of proportions with continuity correction

```{r}
trial %>%
  select(response, death, trt) %>%
  tbl_summary(by = trt, missing = "no") %>%
  add_p(
    test = everything() ~ "prop.test", 
    pvalue_fun = ~style_pvalue(.x, digits = 3)) %>%
  add_n() 

stats::prop.test(table(trial$trt, trial$response))
stats::prop.test(table(trial$trt, trial$death))
```




# Individual models

## Linear regression

```{r}
m <- glm(age ~ trt + grade, trial, family = gaussian)

t1 <- tbl_regression(m) %>%
  add_glance_table()

t1

```

## Logistic regression

```{r}
m <- arm::bayesglm(response ~ trt + age + grade, trial, family = binomial)

t2 <- tbl_regression(m, exponentiate = TRUE) %>% 
  add_vif()

t2
```


## Cox Proportional Hazards Regression

```{r}
library(survival)
cox_m <- coxph(Surv(ttdeath, death) ~ age + marker + grade, data = trial)
tbl_regression(cox_m, exponentiate = TRUE)
```

Note the sensible defaults with this basic usage (that can be customized later):

- The model was recognized as logistic regression with coefficients exponentiated, so the header displayed “OR” for odds ratio.

- Variable types are automatically detected and reference rows are added for categorical variables.

- Model estimates and confidence intervals are rounded and formatted.

- Because the variables in the data set were labelled, the labels were carried through into the {gtsummary} output table. Had the data not been labelled, the default is to display the variable name.

- Variable levels are indented and footnotes added.

### Univariate Cox models

```{r}
cox_uv_m <- trial %>%
  select(ttdeath, death, age, marker, grade) %>%
  tbl_uvregression(
    method = coxph,
    y = Surv(ttdeath, death), 
    exponentiate = TRUE,
    hide_n = TRUE
  ) %>%
  add_nevent(location = "level")

cox_uv_m
```

```{r}
uni_multi <- tbl_merge(
    tbls        = list(
      tbl_summary(
        trial %>%
          select(age, marker, grade)), 
      cox_uv_m, 
      tbl_regression(cox_m, exponentiate = TRUE)),
    tab_spanner = c("**Describe**", "**Univariate Models**", "**Multivariate Model**")
  )

uni_multi
```

## Mixed effects models

```{r}
lme4::lmer(age ~ marker + stage + (1|grade), trial) %>%
  tbl_regression()
```

## Multinomial models

```{r}
library(nnet)
m <- multinom(grade ~ age, trial)

tbl_regression(m)
```


## emmeans - or sort of - simply pairwise tests


```{r}
# table summarizing data with no p-values
small_trial <- trial %>% select(grade, age, response)
t0 <- small_trial %>%
  tbl_summary(by = grade, missing = "no") %>%
  modify_header(all_stat_cols() ~ "**{level}**")

library(gt)
# table comparing grade I and II
t1 <- small_trial %>%
  filter(grade %in% c("I", "II")) %>%
  tbl_summary(by = grade, missing = "no") %>%
  add_p() %>%
  modify_header(p.value ~ md("**I vs. II**")) %>%
  # hide summary stat columns
  modify_column_hide(all_stat_cols())

# table comparing grade I and III
t2 <- small_trial %>%
  filter(grade %in% c("I", "III")) %>%
  tbl_summary(by = grade, missing = "no") %>%
  add_p()  %>%
  modify_header(p.value ~ md("**I vs. III**")) %>%
  # hide summary stat columns
  modify_column_hide(all_stat_cols())

# table comparing grade II and III
t3 <- small_trial %>%
  filter(grade %in% c("II", "III")) %>%
  tbl_summary(by = grade, missing = "no") %>%
  add_p()  %>%
  modify_header(p.value ~ md("**II vs. III**")) %>%
  # hide summary stat columns
  modify_column_hide(all_stat_cols())

# merging the 3 tables together, and adding additional gt formatting
tbl_merge(list(t0, t1, t2, t3)) %>%
  modify_spanning_header(
    list(
      all_stat_cols() ~ "**Tumor Grade**",
      starts_with("p.value") ~ "**p-values**"
    )
  )
```



```{r eval=F}
library(broom.helpers)
mod <- glm(response ~ stage + grade, data = trial, family = binomial)

mod %>%
  tidy_plus_plus(
    add_pairwise_contrasts = TRUE, 
    exponentiate = TRUE
  ) %>%
  knitr::kable()
```





# Further readings and references

- Main page of a package with most of the info you found here: https://www.danieldsjoberg.com/gtsummary/index.html

- Cheat sheet: file:///Users/zablotski/Downloads/gtsummary.pdf

- [The R Journal Article](file:///Users/zablotski/Downloads/RJ-2021-053.pdf)

- Here is a very informative video of {gtsummary} creator - Daniel Sjoberg: https://www.youtube.com/watch?v=tANo9E1SYJE&t=2s&ab_channel=DanielSjoberg

---

If you think, I missed something, please comment on it, and I’ll improve this tutorial.

**Thank you for learning!**







