---
title: "R package reviews {ggstatsplot} visualizations with statistical details (in progress)"
description: |
  "A picture is worth a thousand words"! I'd also say that results of a statistical test are worth a hundrends words. The marriage betwenn them is worth 10.000 words. And that's exactly what ggstatsplot does - creating informative beautiful plots with details from statistical tests, which tells you the story obout your data.
author:
  - name: Yury Zablotski
    url: https://yury-zablotski.netlify.app/
date: 01-09-2021
categories:
  - videos
  - statistics
preview: thumbnail.jpg
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 6
    code_download: true
draft: true
bibliography: /Users/zablotski/Documents/library.bib
#csl: american-political-science-association.csl
biblio-style: apalike
link-citations: yes
linkcolor: blue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
library(tidyverse)
library(ggstatsplot)
```


@Patil2021 created the package, which I used on a daily basis. It is maturing at the moment, so that by the moment you see this video, the package might be even better. 

## Advantages and benefits

- easy consistent code
- rich in information with only 3 arguments,
- beautiful and immideatly publishable
- almost no customization, but possible if needed, works perfectly with ggplot and dplyr syntax
- supports tons of statistical methods which are all now finally unified, because they com from different packages
- reduced statistical errors in reporting


```{r}
# stabilize random selection
set.seed(9)

# randomly select salaries of 100 individuals for both professions: IT and non_IT
d <- ISLR::Wage %>% 
  group_by(jobclass) %>%            
  sample_n(10) %>%               # random selection of individuals in each group
  mutate(id  = row_number()) %>%  # mutate means - create new column
  ungroup() %>%                   # ungrouping will be important for "pivot_wider"
  mutate(job = ifelse(jobclass == "1. Industrial", "non IT", "IT")) %>% 
  select(id, wage, job) 
```


## Comparisons between groups

### 2 groups for t-test and Mann-Whitney test

```{r}
ggbetweenstats(
  data = d,
  x    = job,
  y    = wage
)
```

- type argument will give you 4 values: 

### 3 groups 

r - trimmed means ANOVA

## Comparisons within groups

## Histogramm: distribution of numeric variables

## Dotplot: distribution of labeled numeric variables

## Correlation plots between TWO variables

## Correlations between MULTIPLE variables

## Pie charts for categorical data

## Bar charts for categorical data

## Dot-and-whisker plots for regression models and meta-analysis









## Further readings and lookings

- [The presentation of the package by the author himself](https://www.youtube.com/watch?v=_wCo7rnt6NA&ab_channel=UseROslo)