---
title: "How to create a blog or a website in R with {distill} package (in progress!)"
description: |
  If you're not online, you don't exist. A personal webpage or a blog became the business card of the digital century. It shows who are you and what are you capable of. Thus: show, don't tell.
author:
  - name: Yury Zablotski
    url: https://yury-zablotski.netlify.app/
date: 12-26-2020
categories:
  - R & the Web
  - videos
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    code_download: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, eval=FALSE)
```

# Why should we use distill?

Because it's easy, quick and free way to go online today. All you need to create a website or a blog is R, RStudio and Netlify account. Moreover, it requires very little programming. It might not be too fancy, but it's originally thought to make scientific publishing online, so, it can't be too bad ;).

For a more advanced users, you might wanna use continuous deployment via Github, but you do not understand this sentence, you'll be fine too.

# Manual intermittent publishing (deploying)

Why should we do manual instead of automated and continuous first. Well, for non programmers or IT folks, like me, it might be very frustrating and painful to immediately go with the continuous deploy via Github. We all know what is "better" and "more effective", but we turned off from doing something by complexity, we might never do what we want while thinking "it's not my cap of tea". 

If you are comfortable with Github, click on the "Automated continuous publishing (deploying)" on table of contents on your left.

## Install and load "distill" package

```{r eval=FALSE}
install.packages("distill")
```


```{r}
library(distill)
```

## Create your blog manually or with one line of code

Distill is a publication format for scientific and technical writing, native to the web. 

- open RStudio
- File => New Project... => New Directory => Distill Blog
- "Directory name" will be the name of the folder, where all the documents from your blog will be stored
- Choose the directory where this folder supposed to be stored
- Give your blog a name. You'll be able to rename it.
- Click "Create project"

Learn more about using Distill at <https://rstudio.github.io/distill>.





or execute the code below:

The easiest way to get started is to use the Distill Blog option in the RStudio New Project dialog:

```{r eval=FALSE}
create_blog(dir = "my blog", title = "A cool name of my blog")
```


## Knit it

The "knit" button will show you how your website will look like. You can also open it in your browser.

## Publish your website via Netlify

- sing up for Netlify via Email (you'll need to conferm your Email)
- In "Team overview" you'll see a window with: 
"Want to deploy a new site without connecting to Git?
Drag and drop your site folder here"
- Drag and Drop your "_sites" into this window
- wait a bit and you'll be getting a funny named website name, click on it
- Contrats, you are online!
- Go to "Site settings" => Change site name => choose a "better-name" => Save
- Click on the "better-name.netlify.app" to make sure the name changed and the site is still working.
- go back to RStudio and create a new post

## Create a new blog-post

```{r eval=FALSE}
distill::create_post("My second blog post...this will be really good :-)")
```

- these new post are not on your local computer, but still not online, thus,
- go back to Netlify => Deploys (tab) where you'll see: "Need to update your site? Drag and drop your site folder here" 
- drag and drop the "_site" folder there and wait till "Production" (button) terns green and says Published.
- Click on the "better-name.netlify.app" to make sure the new blog-post appeared. 




# Automated continuous publishing (deploying)

## Create blog either manually or with one line of code

- open RStudio
- File => New Project... => New Directory => Distill Blog
- "Directory name" will be the name of the folder, where all the documents from your blog will be stored
- Choose the directory where this folder supposed to be stored
- Give your blog a name. You'll be able to rename it.
- Click "Create project"

or execute the code below:

```{r}
create_blog(dir = "my blog", title = "A cool name of my blog")
```

## Have a look at your blog

Either Knit the document, or go to the "Build" tab and press "Build Website".

## Connect to your github

You have to have "git" installed on your computer and have a github profile. If you still aren't, [Happy Git and GitHub for the useR](https://happygitwithr.com/) will help you with that. Take your time with this short, but comprehensive book.

### Then install "usethis" package, if you need:

```{r eval=FALSE}
install.packages("usethis")
```



### Run use_git() and answer two questions:

- Run "use_git()" command. You'll be asked: "Is it ok to commit them?"
- Do not commit yet. Type 3 for "No", or "Nope" or similar. You'll get another question: "A restart of RStudio is required to activate the Git pane
Restart now?"
- Do restart! Type 2 for "Yes" or "Yup".

```{r eval=FALSE}
usethis::use_git() 
```


After restart you'll see a "Git" tab near the Environment, History etc. That's gut! ;) Click on the "Git" tab and you'll see Empty boxed under "Staged", lot's of yellow question marks under "Status" and the documents names under "Path". 

### Stage and Commit

- Check all the boxes and press "Commit" button, which is between the "Status" and "History" tabs. A colorful window will pop up. This window will describe all the changes you are about to make to your blog. 

- Definitely describe what you have done for a future you, because then you'll always be able to get back to this, in case something stops working. That's what they call - a version control. 

- Write something like "first commit" and press "Commit". 

- Close or forget the window and go back to RStudio.


### Run use_github() to create a new Github repository


```{r eval=FALSE}
usethis::use_github() 
```

A new repository will be automatically created on your Github profile and the new browser window with your Github will pop up.

If you'll be asked: "Which git protocol to use?", choose the one with "https" and if you'll then be asked: "Are title and description ok?", agree with "Yes" or "Ok" or similar.

NOTE: if something else doesn't work, get back to the "Happy Git" book and work through it. You'll only need it once. Three main things you'd need to get working are:

- install git
- connect RStudio to your Github and 
- configure a GitHub Personal Access Token

## Publish your blog with Netlify

If you have a Notlify account you are fine, if not you can create one using your Github account. Then press:

- "New site from Git" in a green box

- Click on Configure Netlify on "GitHub", 

- Choose a newly created repository, you'll recognise the name ("my blog"?).

- **NOTE:** Make sure to set the "Publish Directory" to "_site" (could be "docs" if you checked some boxes while creating new project)

- Click deploy!

- you'll get an absurd or funny random name for your website (e.g. gracious-mestorf-194a9f), which you immediatly by 

- clicking "Site settings" => "Change site name"

- now go on and click on the link of your blog. Congrats! Your blog is online! ;)

- go back to RStudio and create a new post

## Create a new blog-post

- run following command in the Console:

```{r eval=FALSE}
create_post("My second blog post...this will be really good :-)")
```

- a new RMarkdonw document will be created and opened. 
- these new post is only on your local computer, but still not online, thus,
- "Knit" the document 
- go to the "Git" tab and check all the boxes
- press "Commit"
- add some description
- again press "Commit"
- wait until you see the "Close" button and close "Git commit" pop up window
- press "Push"
- wait until you see the "Close" button and close "Git push" pop up window
- close or ignore the other pop up window.
- Get back to your Webpage and refresh (it may take a few seconds, so, don't panic if the first refresh don't work). You should see a new blog-post.

Now you can repeat following routine for every post:
- Create or change posts
- Knit
- Commit
- Push

Congrats! You are continuosly deploy your online blog. 

## Always Knit after changing

By the way, if you don't want your post to publish until you really sure, you can start out as a draft:

```{r eval=FALSE}
distill::create_post("Another nice post", draft = TRUE)
```

"Knitting" blog posts is the only way to update it. They are not re-rendered when the site is built, because it's computationally expensive and prone to problems. Build Website would not work for blogs. 




# Blog configuration with "_site.yml"

- Open "_site.yml". You'll see something like that:

```{r}
name: "reproducible-finance-with-r"
title: "Reproducible Finance with R"
description: |
  Exploring reproducible finance with the R statistical 
  computing environment.
base_url: https://beta.rstudioconnect.com/content/11424/
navbar:
  logo: images/rstudio.png
  right:
    - text: "Home"
      href: index.html
    - text: "About"
      href: about.html
    - text: "Contributors"
      href: contributors.html
    - icon: fa fa-rss
      href: index.xml
output: distill::distill_article
```




# Improve your website or blog

- Listing pages

- Categories
 
Categories can be added similar to the blogdown in the 


  
```{r}
title: "How to create a blog or a website in R with {distill} package (in progress!)"
author:
  - name: Yury Zablotski
date: 12-26-2020
categories:
  - R & the Web
  - videos
```

- Thumbnails

Thumbnails (or previews) are the images which are displayed along the post. They serve as an eyecatcher. You can add a preview image into the post’s metadata by adding a "preview" field:

```{r}
title: "Blog post on how to write blog posts"
description: |
  Here we were out of ideas, and therefore we are proud to announce
our new post about how to write a post. Creativity is a b**ch! ... And I love her!
preview: images/photo-of-me-because-I-am-sooo-beautifuuulll-toniiiight.png
```

  
  
# Theming - change desing and appearence of your blog


Distill blogs can use themes and CSS just like Distill sites. You can use the create_theme() function to add a theme CSS file in the current working directory:

create_theme(name = "theme") 

```{r}
distill::create_theme(name = "theme") 
```

To apply a custom theme site-wide, add a theme key to the top-level of your "_site.yml" configuration file:

```{r}
name: "distill"
title: "Distill for R Markdown"
theme: theme.css 
```



# Difference between a website and a blog

All articles on a website are re-rendered when the site is rebuilt while the blog-post are not. We need to knit each of the article separately. This might sound bad first, because we are not uptodate with our blog, but I personally found it great. See, like the authors of "distill" state: "Given that R package upgrades have a tendency to break older code, continuously re-rendering old posts is nearly impossible to do without errors, especially over longer periods of time." That happend to my old blogdown-generated website, where Hugo-Academic theme changes so dramatically that I could not modify my site anymore, and, since it was not on Github, I could not even ask R community, what is wrong. Thus, I totally understand, if you go for a manual deployment first (been there!), but I would recommend you to progress to Github and continuous deploy after you have got comfortable. Especially if you create a lot, automated deployment will save you lot's of time. However, my old blogdown-website is still online and running. Thus, I prefer blog instead of the website.


# Useful ressources

- Tom Mocks blog-post on how to build blogs with distill: https://themockup.blog/posts/2020-08-01-building-a-blog-with-distill/

