---
title: "4 Reasons Non-Parametric Bootstrapped Regression (with tidymodels) is Better thаn Ordinary Regression"
description: |
  If the assumptions of parametric models can be satisfied, parametric models are the way to go. However, there are often many assumptions and to satisfy them all is rarely possible. Data transformation or using non-parametric methods are two solutions for that. In this post we'll learn the Non-Parametric Bootstrapped Regression as an alternative for the Ordinary Linear Regression in case when assumptions are violated.
author:
  - name: Yury Zablotski
    url: https://yuzar-blog.netlify.app/
date: "`r format(Sys.time(), '%B %d, %Y')`"
categories:
  - videos
  - statistics
  - visualization
  - models
preview: thumbnail_bootstrapped_regression.png
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 6
    code_download: true
bibliography: /Users/zablotski/Documents/library.bib
#csl: american-political-science-association.csl
biblio-style: apalike
link-citations: yes
linkcolor: blue
# draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```


# This post as a video (in making)

I recommend to watch a video first, because I highlight things I talk about. It's less then ... minutes long.

```{r, eval=T, echo=F}
vembedr::embed_youtube("sjCxIHVGkdE") 
```

# The problem with ordinary linear models

If we want to compare salaries of people with different education, we could create a simple linear model and plot the results. But we can not trust these results, without checking the assumptions of our model, because if assumptions are violated, the results might be wrong. However, the problem with assumptions is that there are too many, and most of the time we can't satisfy them all. 


```{r}
library(tidymodels)   # for everything good in R ;)
library(ISLR)         # for "Wage" dataset

set.seed(9999)        # makes sure we get the same 100 lines
salary <- Wage %>%
  sample_n(100)

m <- lm(wage ~ education, salary)

library(sjPlot)      # I have a video on "sjPlot" 📦
theme_set(theme_test())
plot_model(m, type = "pred", show.data = TRUE, jitter = TRUE)
```


For instance, our model has not-normally distributed residuals, the variances between groups differ and we have heteroskedasticity. So, we can't trust our results, without fixing them, but when we start fixing assumptions, we might (1) either lose data (if, for example, we remove some influential points), (2) we might reduce interpretability (when we log-transform data) or screw up other assumptions even more, because data manipulation constantly changes the **parameters** of our - **parametric** model. And here is where **non-parametric bootstraped regression** comes into play. 


```{r}
library(performance) # I have a video on "performance" 📦 

check_normality(m)          # Shapiro-Wilk normality test
check_normality(m) %>% plot(type = "qq")

check_homogeneity(m)        # Bartlett test 
check_homogeneity(m) %>% plot()

check_heteroscedasticity(m) # Breusch-Pagan test
check_heteroscedasticity(m) %>% plot()
```




# Why should we nonparametrically bootstrap regressions?





The 4 reasons a bootstrapped model is better then usual linear model, is that bootstrapping:

1. does not have any distributional assumptions (such as normally distributed residuals or equal variances among groups)

2. bootstrapped models provide more accurate inferences, e.g. confidence intervals [^1], which we'll soon prove on these two examples, one with categorical and one with numeric predictors

3. bootstrapped models work better for small samples and finally 

4. it describes variation in the data much better than traditional linear models

Therefore, having bootstrapped models in your tool-box will definitely make you a better data-scientist! So ...

# What is - a bootstrap?

In order to better understand how bootstrap models work, we need to understand what the bootstrap itself actually is. A bootstrapp is the little loop on the back of a boot to help you pool it on. And the phrase "pulling oneself up by one’s bootstraps" means to succeed without any external help. For data it simply means - resampling.

```{r}
library(ggstatsplot) # I made many videos on "ggstatsplot" 📦
ggbetweenstats(data = salary, x = education, y = wage)
```

So, lets (1) take a very small dataset, (2) resample it 1000 times, (3) calculate bootstrapped coefficients & 95% CIs and (4) compare them to the ordinary regression in order to find out whether bootstrapping model is indeed better!

# 4 steps to compute bootstrapped regression

There are only 4 steps to conduct and visualize bootstrapped regression. 

## 1. Bootstrap the data with resampling

First, we bootstrap the data, which simply means we take 1000 samples from our 100 data rows. Every of those 1000 samples is of the same size as the original data set, namely 100 rows, but is made using **replacements**, which results into multiple replicates of some of the original rows of the data. **Replacements are necessary, because we would otherwise simply reproduce the original sample**. (The assessment set contains the rows of the original data that were not included in the bootstrap sample.) In order to distinguish bootstrapped samples from the original sample let's call our bootstrapped samples - splits. 


```{r}
set.seed(123)
# 1000 samples are OK, but with 2000-10000 you are save!
boot_data <- bootstraps(salary, times = 1000) 
boot_data
boot_data$splits[[1]]
```

## 2. Produce 1000 models

Then we **fit a linear model** to every **split** using the first `map()` function and tidy up model coefficients with the second `map()` function. Let's use "0 +" in the model formula to remove the Intercept, because we are interested in estimates of the salary and not the slope of change from that Intercept.

```{r}
boot_models <- boot_data %>% 
  mutate(model = map(splits, ~lm(wage ~ 0 + education, data = .)), 
         coefs = map(model, tidy) ) # optional "conf.int = TRUE"

boot_models
```


## 3. Plot estimates distribution of 1000 estimates

```{r}
boot_coefs <- boot_models %>%
  unnest(coefs) %>% 
  select(term, estimate)

boot_coefs
```

Now, we not only have 1000 models, which are nested in the column "model", but also the results of those models which are nested in the column "coefficients". And if we `unnest()` the coefficients, we'll see 5 estimates for every of our 1000 models. It's like we have done 1000 experiments, which we can immediately visualize as a distribution.

```{r fig.width=9}
boot_coefs %>%
  ggplot(aes(x = estimate)) +
  geom_histogram() +
  facet_wrap(~term, scales = "free")
```
First of all, the distributions seem normal, or bell-shaped, which is already amazing because we can use average estimates ;). But despite normal distribution, I still would prefer median estimate for every group instead of the mean. That makes our non-parametric estimates even more robust, since median is estimated by another wonderful robust technique - **quantile-regression**, which is also used when the assumptions of linear regression are not met. I find it really cool! But the coolest thing about this distribution and quantile-function is that we can easily get not only classic 95% Confidence Intervals, **but any intervals we want**. Which is kind of hard to get out of the ordinary model. Thus, we actually have more than 4 advantages 🥳 of the bootstrapped model. 

```{r}
boot_meds <- boot_coefs %>% 
  group_by(term) %>% 
  summarise(
    med_est   = median(estimate),
    quantile  = quantile(estimate, 0.5  ),
    conf.low  = quantile(estimate, 0.025),
    conf.high = quantile(estimate, 0.975),
    conf.25   = quantile(estimate, 0.25 ),
    conf.75   = quantile(estimate, 0.75 ))

boot_meds
```

And when we plot our estimates we can clearly see that 50% of people who did not finish a high school, will never reach a salary of 100 thousands, while 95% of folks with higher education will never earn below 100 thousands bugs. Waw! So, education matters!

```{r fig.width=9}
boot_coefs %>%
  ggplot(aes(x = estimate)) +
  geom_rect(aes(x = med_est, xmin = conf.low, xmax = conf.high, ymin = 0, ymax = Inf), data = boot_meds, alpha = 0.1, fill = "green") +
  geom_rect(aes(x = med_est, xmin = conf.25, xmax = conf.75, ymin = 0, ymax = Inf), data = boot_meds, alpha = 0.3, fill = "green") +
  geom_density() +
  geom_vline(data = boot_meds, aes(xintercept = med_est))+
  facet_wrap(~term, scales = "free")
```

But what blew my mind even more the first time I learned to bootstrap regression estimates, is that I can even get a distribution of p-values :). For that we just (1) remodel 1000 ... ahhh, what a hell, let's remodel 10.000 models with the Intercept, (2) unnest out coefficients again and (3) visualize the distribution of our 10.000 p.values. Now you see why it's always better to take median instead of mean. 


```{r fig.width=9}
boot_models <- bootstraps(salary, times = 10000) %>% 
  mutate(model = map(splits, ~lm(wage ~ education, data = .)), 
         coefs = map(model, tidy) ) # optional "conf.int = TRUE"

boot_coefs <- boot_models %>%
  unnest(coefs) %>% 
  select(term, p.value)

boot_coefs %>%
  ggplot(aes(x = p.value)) +
  geom_histogram() +
  facet_wrap(~term, scales = "free")
```

And if we compare mean and median bootstrapped p.values to the p.values from the ordinary models, we'll see that median bootstrapped p.values are much closer to the the p.values of the ordinary model. And despite the fact that median p.values and ordinary p.values are only slightly different, I would still intuitively trust the bootstrapped p.values more, because they were produced in a statistically robust fashion, where no assumptions were violated.

```{r}
boot_coefs %>% 
  group_by(term) %>% 
  summarise(
    mean_boot_p = mean(p.value),
    med_boot_p  = median(p.value)
    ) %>% 
  # remember? m <- lm(wage ~ education, salary)
  left_join(tidy(m) %>% select(term, p.value)) %>% 
  mutate_if(is.numeric, ~round(., 4))
```


## 4. Get and visualize predictions

Finally, here is the moment we've been waiting for: 

- let's use the `augment()` command to produce predictions for every of our 10.000 models, 
- then calculate a median and 95% CIs from our predictions
- calculate predictions made by the ordinary linear regression
- combine two results into one dataset and
- compare them visually by 
  - first plotting our original 100 observations and then
  - display predictions of both models to see which describes the data best

```{r}
# get predictions
boot_aug <- boot_models %>% 
  mutate(augmented = map(model, augment)) %>% 
  unnest(augmented) %>% 
  select(-splits, -model)

# get median estimates, median 95% CIs
nonpar_med_boot_preds <- boot_aug %>% 
  group_by(education) %>% 
  summarise(
    predicted = median(.fitted ),
    conf.low  = quantile(.fitted, 0.025),
    conf.high = quantile(.fitted, 0.975)) %>% 
  mutate(model = "bootstrapped") %>% 
  select(model, everything())

nonpar_med_boot_preds

# get mean estimates from the classic model
m <- lm(wage ~ education, salary)  # just a reminder
par_avg_preds <- ggeffects::ggeffect(m, terms = "education") %>% 
  tibble() %>% 
  mutate(model = "linear") %>% 
  select(model, education = x, predicted, conf.low, conf.high)

par_avg_preds

# combine two datasets
preds <- rbind(nonpar_med_boot_preds, par_avg_preds)

# plot row data and predicted CIs from both models
ggplot() +
  geom_jitter(data = salary, aes(education, wage), 
              width = .2, alpha = 0.2, size = 2)+
  geom_pointrange(data = preds, aes(x = education, y = predicted, 
                  ymin = conf.low, ymax = conf.high, color = model),
                  position=position_dodge(width=.6), size = 1)
```

<!-- Interestingly, the center of the bootstrap distribution is identical to the mean of the sample, because the bootstrap distributions are centered around statistics calculated from the data (e.g., a regression slope). However,  -->

Interestingly, the bootstrapped confidence intervals are slimmer where the variance is low and wider, where the variance is high. So, as mentioned in the beginning, **bootstrapped models describe the variation in the data better and produce therefore more accurate and realistic inferences without violating any assumptions, especially for small samples.**

# Bootstrapping numeric predictors

Bootstrapping with numeric predictors works in the same way, so, let's summarise all we have learned so far, namely: 

- we first create 1000 resamples of our data,
- then `map()` over it to fit new 1000 models, with the numeric predictor "age"
- we then `map()` over every model and use `augment()` function to extract fitted data for every model in an new nested column, with 100 fitted values for every model, because every resample in boot_data has 100 observations.
- 1000 models with 100 fitted values each result in 100.000 fitted values, which we see when we unnest them for plotting, and visualize as 1000 lines on the plot


```{r}
# produce 1000 resamples
set.seed(123)
boot_data <- bootstraps(salary, times = 1000) 

# bootstrap new regressions and unnest all fits
boot_models <- boot_data %>% 
  mutate(model     = map(splits, ~lm(wage ~ age, data = .)), 
         augmented = map(model, augment) )

boot_models

boot_aug <- boot_models %>% 
  unnest(augmented) %>% 
  select(-splits, -model)

boot_aug

ggplot(data = boot_aug, aes(x = age, y = wage)) +
  geom_line(aes(y = .fitted, group = id), col = "grey",  size = 0.25) 
```

- we then create **new age** data (picture of new age movement) ... ups, sorry, not that 
**new age**, just data of age from 20 to 75 years old, and call them new in order to evaluate our predictions
- use `map()` function to predict salaries for the new age data by every model
- then summarize all 1000 predictions by the median and calculate 95 and 50% confidence intervals (CIs),
- and finally plot them on top our our 1000 fitted models

```{r}
# get new data for predictions
new <- data.frame(age = seq(20, by = 3, 75))
new %>% t()

few_preds <- boot_models %>% 
  mutate(few_preds = map(model, predict, new)) %>% 
  unnest(few_preds) %>% 
  select(-splits, -model) %>% 
  mutate(age = rep( seq(20, by = 3, 75) , 1000)) %>% 
  group_by(age) %>% 
  summarise(wage   = median(few_preds),
            LCL_50 = quantile(few_preds, 0.25),
            UCL_50 = quantile(few_preds, 0.775),
            LCL    = quantile(few_preds, 0.025),
            UCL    = quantile(few_preds, 0.975),
  )

few_preds

ggplot(data = boot_aug, aes(x = age, y = wage)) +
  geom_line(aes(y = .fitted, group = id), col = "grey",  size = 0.25) +
  geom_point(data = few_preds, aes(age, wage))+
  geom_errorbar(data = few_preds, aes(ymin = LCL, ymax = UCL), width = 1)+
  geom_errorbar(data = few_preds, aes(ymin = LCL_50, ymax = UCL_50))
```

- then we'll plot the original 100 data points in green and
- the results of an ordinary linear regression in blue with their 95% CIs in red 

```{r}
ggplot(data = boot_aug, aes(x = age, y = wage)) +
  geom_point(data = salary, aes(age, wage), colour = "green", shape = 1) +
  geom_line(aes(y = .fitted, group = id), col = "grey",  size = 0.25) +
  geom_point(data = few_preds, aes(age, wage))+
  geom_errorbar(data = few_preds, aes(ymin = LCL, ymax = UCL), width = 1)+
  geom_errorbar(data = few_preds, aes(ymin = LCL_50, ymax = UCL_50))+
  geom_smooth(aes(x = age, y = wage), data = salary, method = "lm",
              colour = "blue", fill = "red", alpha = 0.25)


ggsave("num_example.jpeg", plot = last_plot(), width = 7, height = 4)
```


And as we can see again, 

- low variance in salaries in the beginning of professional life is described with narrower CIs by the bootstrapped predictions as compared to the ordinary linear regression, while 
- larger variance in salaries after age of 45 is described by the bootstrapped predictions with wider CIs as compared to the ordinary linear regression

Thus, similarly to the categorical predictor, we see that **bootstrapped results better describe the variation in the data and produce therefore more accurate and realistic inferences.**


# Problems

So, is bootstrap method perfect? Of coarse no! While bootstrap more accurately describes the variance of a sample, for example here I really want to include high salaries into my variance, to see what is actually possible to earn with this education, if you have real outliers or very influential observations in your data, they will be given more weight then needed. In this case you would use a **robust regression**, and if you wanna become a more complete data scientist in the next five minutes, watch [this video](https://youtu.be/M_7MOkAm9WU).  


```{r}
library(robustbase)
rm <- lmrob(wage ~ education, salary)
rob_avg_preds <- ggeffects::ggeffect(rm, terms = "education") %>% 
  tibble() %>% 
  mutate(model = "robust") %>% 
  select(model, education = x, predicted, conf.low, conf.high)

# combine two datasets
preds <- rbind(nonpar_med_boot_preds, par_avg_preds, rob_avg_preds)

ggplot() +
  geom_jitter(data = salary, aes(education, wage), width = .2, alpha = 0.2, size = 2)+
  geom_pointrange(data = preds, aes(x=education, y=predicted, 
                                    ymin=conf.low, ymax=conf.high, color = model),  
                  position=position_dodge(width=.6), size = 1)

ggsave("rob_example.jpeg", plot = last_plot(), width = 7, height = 4)
```


# Further readings and references

- https://www.datawim.com/post/bootstrapping-regression-coefficients-in-r/

- https://padpadpadpad.github.io/post/bootstrapping-non-linear-regressions-with-purrr/

[^1]: Tim Hesterberg (2015), What Teachers Should Know about the Bootstrap: Resampling in the Undergraduate Statistics Curriculum, The American Statistician 69(4) 371-386, DOI: 10.1080/00031305.2015.1089789


---

If you think, I missed something, please comment on it, and I’ll improve this tutorial.

**Thank you for learning!**
