---
title: "R package reviews {gtsummary} Effective Table Creation!"
description: |
  A short description of the post.
author:
  - name: Yury Zablotski
    url: https://yuzar-blog.netlify.app/
date: "`r format(Sys.time(), '%B %d, %Y')`"
categories:
  - videos
  - statistics
  - models
  - machine learning
preview: thumbnail_bootstrapped_regression.png
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 6
    code_download: true
bibliography: /Users/zablotski/Documents/library.bib
#csl: american-political-science-association.csl
biblio-style: apalike
link-citations: yes
linkcolor: blue
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```

# This post as a video (in making)

I recommend to watch a video first, because I highlight things I talk about. It's less then ... minutes long.

```{r, eval=T, echo=F}
vembedr::embed_youtube("") 
```

# Simple descriptive table

## tbl_summary()

```{r}
library(gtsummary)

sm_trial <- 
  trial %>% 
  select(trt, age, grade, response)

sm_trial %>% 
  select(-trt) %>% 
  tbl_summary()
```

```{r}
trial2 <- trial %>% select(trt, age, grade)
trial2 %>% tbl_summary()
```


## include statistical tests

```{r}
trial2 %>% tbl_summary(by = trt) %>% add_p()

iris %>% tbl_summary(by = Species) %>% add_p()
```

## Add function to tbl_summary

- add_*()
- modify_*()
- bold_* () / italicize_* ()

```{r}
library(ISLR)

Wage %>% 
  select(wage,education, jobclass, health) %>% 
  tbl_summary(by = jobclass) %>% 
  add_p() %>% 
  add_q(method = "BH")
```
```{r}
mtcars %>% 
  select(am, mpg, cyl) %>% 
  tbl_summary(by = am) %>% 
  add_difference()
```

```{r}
mtcars %>% 
  select(am, mpg, cyl) %>% 
  tbl_summary(by = am) %>% 
  add_p() %>% 
  add_overall() %>% 
  add_n() %>% 
  add_stat_label()
```

## inline_text()

## tbl_merge() & tbl_stack()

## themes()

```{r}
theme_gtsummary_compact()

library(ISLR)    # for Auto dataset
m <- lm(mpg ~ origin * horsepower, data = Auto)
tbl_regression(m) %>% 
  modify_caption("Compact theme")
```


# Further readings and references

- Here is an amazing and informative video of {gtsummary} creator - Daniel Sjoberg: https://www.youtube.com/watch?v=tANo9E1SYJE&t=2s&ab_channel=DanielSjoberg

---

If you think, I missed something, please comment on it, and I’ll improve this tutorial.

**Thank you for learning!**







