---
title: "R demo | Cochran Test + Pairwise McNemar Tests (post-hoc)"
description: |
  A short description of the post.
author:
  - name: Yury Zablotski
    url: https://yuzar-blog.netlify.app/
date: "`r format(Sys.time(), '%B %d, %Y')`"
categories:
  - videos
  - statistics
preview: thumbnail.png
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 6
    code_download: true
bibliography: /Users/zablotski/Documents/library.bib
#csl: american-political-science-association.csl
biblio-style: apalike
link-citations: yes
linkcolor: blue
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```


## This post as a video

I recommend to watch a video first, because I highlight things I talk about. It's ca. ... minutes long. 

```{r, eval=T, echo=F, fig.height=5, fig.width=7}
vembedr::embed_youtube("")
# library(tidyverse)
```


## Get the data

```{r}
set.seed(9) # for reproducibility 
data_wide <- data.frame(
  before    = sample(c("Positive",
                       "Negative",
                       "Positive"), 30, replace = TRUE),
  after     = sample(c("Negative",
                       "Positive",
                       "Negative"), 30, replace = TRUE),
  longtirm  = sample(c("Negative",
                       "Negative",
                       "Positive"), 30, replace = TRUE))
```


## Compute Cochran & Pairwise McNemar Tests

```{r}
# install.packages("tidyverse")
library(tidyverse)

data_long <- data_wide %>% 
  mutate(id = 1:nrow(.)) %>% 
  gather(key = "time", value = "outcome", before:longtirm) %>% 
  mutate_all(factor)

View(data_long)

# test for comparing more than two paired proportions.
# install.packages(rstatix)
library(rstatix)
mcnemar_test(data_wide$before, data_wide$after, correct = F)

xtabs(~outcome + time, data_long)

# pairwise_mcnemar_test: performs pairwise McNemar's chi-squared test 
# between multiple groups. Could be used for post-hoc tests following a 
# significant Cochran's Q test.
cochran_qtest(data_long, outcome ~ time|id)
pairwise_mcnemar_test(data    = data_long, 
                      formula = outcome ~ time|id, 
                      correct = F, 
                      p.adjust.method = "holm")
```



## Interpret the result


**So, the final interpretation of our Cochran Test + Post-Hoc McNemar tests is: .** Which is amazing, since we can stop aliens. However, if we took a wrong test the things could go really wrong. Let me show you what I mean.


## What happens if we choose a wrong test





## What's the next best thing to learn?

- Paired t-test or paired wilcoxon test
- Mixed-Effect models

---

If you think, I missed something, please comment on it, and I’ll improve this tutorial.

**Thank you for learning!**



## Reference










